---
title: "PARASITE ANALYSIS"
author: "Tamar Feldman"
date: '2023-08-13'
output: html_document
---

<font size="4">Project: Feldman et al. _Blood Adv_. 2023  
Purpose:  
- Compare parasite gene expression at different host cell stages to blood stage transcriptome.  
- Tally transcripts encoding known or suspected exported effectors  
Required Files:  
- Pf_Published_Otto16.csv  
- PfGeneCountsTable_v3.csv  
- ParasiteColData.csv  
- PlasmodiumExportomeEdited.csv
</font>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIBRARIES

```{r include = FALSE, eval=FALSE}
library(dplyr)
library(stringr)
library(tidyr)
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(grid)
```

# PREPARE FILES
## OTTO ET AL
Load the blood stage transcriptome published by Otto et al.
```{r eval=FALSE}
otto.file.path <- "../REFERENCE_FILES/DIFFERENTIAL_EXPRESSION/Pf_Published_Otto16.csv"
otto.counts <- read.csv(otto.file.path, header = T)
otto.counts <- data.frame(otto.counts, row.names = "gene_ids")
#%>%
#  filter(rowMeans(.) >= 5)
print(head(otto.counts))
```

## MY COUNTS
Load the table of parasite gene counts from this study.
```{r eval=FALSE}
my.file.path <- "../PROCESSED_DATA/ALIGNMENT_RESULTS/PfGeneCountsTable_v3.csv"
my.counts <- read.csv(my.file.path, header = T)
rownames(my.counts) <- my.counts$Gene.ID
my.counts <- my.counts[,2:ncol(my.counts)]
head(my.counts)
```
# COMPARE BLOOD STAGE AND ERYTHROBLAST

## DESEQ2 FOR NORMALIZED COUNTS
Use DESeq2 to produce a table of normalized counts
```{r eval=FALSE}
my.col.path <- "../REFERENCE_FILES/DIFFERENTIAL_EXPRESSION/ParasiteColData.csv"
my.coldata <- read.csv(file = my.col.path, header = T)
my.coldata$cell.type <- factor(my.coldata$cell.type,levels = c("ProE","BasoE","PolyE","OrthoE"))
my.dds <- DESeqDataSetFromMatrix(countData = my.counts, colData = my.coldata, design = ~cell.type)
keep <- rowSums(counts(my.dds) >= 10) >= 4
my.dds <- my.dds[keep,]
my.dds <- estimateSizeFactors(my.dds)
my.pf.counts <- counts(my.dds, normalized = T)
```

## MERGE
Merge the counts from this study and Otto et al. into a single table
```{r eval=FALSE}
all.counts <- merge(my.pf.counts,otto.counts, by = 0)
print(head(all.counts))
```

## SPEARMANN
Use Spearmann's correlation for comparison (see Ludwig et al. Cell Rep. 2019 comparison of erythroblast transcriptomes with published studies)
```{r eval=FALSE}
cormat <- cor(x = all.counts[,2:18], y = all.counts[,19:length(colnames(all.counts))], method = "spearman")
cormat2 <- cor(all.counts[,-1],method = "spearman")
```

# EXPRESSION OF EXPORTED EFFECTORS
Load list of parasite "exportome" genes and select genes expressed at the troph stage according to Otto et al. 
```{r eval=FALSE}
export.file.path <- "../REFERENCE_FILES/DIFFERENTIAL_EXPRESSION/PlasmodiumExportomeEdited.csv"
export.genes <- read.csv(export.file.path, header = T)
# get non-zero counts at either stage
troph.genes <- otto.counts %>%
  select(hrs_8, hrs_16) %>%
  filter(rowSums(.) != 0)
export.troph <- merge(troph.genes, export.genes, by.x = 0,by.y =
"PlasmoDB.ID")
```

```{r eval=FALSE}
export.troph <- export.troph[!duplicated(export.troph$Row.names),]
# merge with expressed genes
my.export <- merge(my.pf.counts, export.troph,by.x = 0, by.y = "Row.names")
print(head(my.export))
my.export <- my.export[!duplicated(my.export$Row.names),]

write.csv(x = my.export, file = "../PROCESSED_DATA/DESEQ2_RESULTS/PARASITE/ExportedOverlap_Output.csv")
```